EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces

### 2018-11-08 BrainScience/brainscience.tex
Write-up active and passive modalities

\begin{itemize}
\item Attenuation: Also called suppression or depression. Reduction of amplitude of EEG activity resulting from decreased voltage. When activity is attenuated by stimulation, it is said to have been "blocked" or to show "blocking".
\item Hypersynchrony: Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range. The term suggest an increase in the number of neural elements contributing to the rhythm, or in the synchronization of different neurons with the same discharge pattern~\citep{Buzsaki2012}.
\item Paroxysmal: Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity. 
\item Monomorphic: Activity appearing to be composed of one dominant waveform pattern.
\item Polymorphic: Activity composed of multiple frequencies that combine to form a complex waveform.
\item Transient: An isolated wave or pattern that is distinctly different from background activity. 
\end{itemize}

@@

The traditional clinical approach consists in analyzing the paper strip that is generated by the plot of the signal obtained from the device.  Expert technician and physicians analyze visually the plots looking for specific patterns that may give a hint of the underlying cognitive process or pathology.   Atlases and guidelines were created in order to help in the recognition of these complex patterns.   Even Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .  The clinical EEG research has also focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements} \citep{Schomer2010}.

Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)~\citep{Boostani2017,Rodenbeck2006}, where the different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines~\citep{Dimitriadis2018}. Visual characterization includes the identification or classification of certain waveform components, or transient events, based on a subjective characterization (e.g. positive or negative peak polarity) or the location within the strip.  It is regular to establish an amplitude difference between different waveforms from which a relation between them is reckoned and a structured index is created (e.g. sleep K-Complex is well characterized based on rates between positive vs negative amplitude) \citep{Uchida1999}.  Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,  sleep spindles, polysplindles, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).

Moreover, EEG data acquisition is a key procedure during the assessment of patients with focal Epilepsy for potential seizure surgery, where the source of the seizure activity must be reliably identified. The onset of the Epileptic Seizure is defined as the first electrical change seen in the EEG rhythm which can be visually identified from the context and it is verified against any clinical sign indicating seizure onset.  The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges\citep{EEGIntro}.

Waveform characterization is the method in which analysis has been performed for Event Related Potentials.  Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.  ERPs are regularly used to assess auditory response in infants.  They are extensively used and studied in Cognitive Neuroscience~\citep{Luck2005}.  ERPs are identified by their components which are recognizable signal shapes assigned to the observed waveform, that can be linked to some cognitive or measurable psychological process.   The P300 is one of the most studied ERP component in BCI because it can be used to implement a Speller application.  Hence, P300 ERPs are a target phenomena to study by automatic waveform recognition methods.

Table~\ref{tab:methods} shows a list of depictions used to describe waveforms in the surveyed literature.  Epilepsy has been described by the nature of oscillatory characterization of their waves, like ripples and wiggles, imitating sawtooths or by their geometric shape.  For ERPs on the other hand, more elaborate indexes has been provided, establishing relations between amplitudes of signal components.  Finally, Sleep studies and ICU research are areas where the most complex indexes has been derived, particularly the coupling of signal properties like phase, amplitude and frequency.

@@

The search of methods based on waveforms is conducted by following the PRISMA~\citep{Moher2009} guidelines.  Search is performed on Google Scholar, Semantic Web and IEEE Xplore search engines by the terms "Waveforms" OR "Shape" OR "Morphology" OR "Visual inspection" + "EEG".

@@

A related method is used in \citep{Alvarado-Gonzalez2016} where the area under the curve of the EEG is sumarized to derive a feature.  This was even used in the seminal work of Farwell and Donchin on P300 \citep{Farwell1988,WolpawJonathanR2012}. Additionally, a logarithmic graph of the peak-to-peak amplitude which is called amplitude integrated EEG (aEEG) \citep{Shah2015} is used nowadays in Neonatal Intensive Care Units.

Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature derived from the amplitude and frequency of its signal and its derivative in time-domain is used.  Moreover, other approaches explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied. Finally the proposals of Burch, Fujimori, Uchida and the Period Amplitude Analysis (PAA) \citep{Uchida1996} algorithm are few of the earliest depictions where the idea of capturing the shape of the signal were established.

@@

Once the plot is generated, its center is used to localize the center of the SIFT patch. This region of the image, where the signal's most important salient waveform should be located, is divided in a grid of $4 x 4$ block and the bidimensional gradient vectors are calculated on each one of them.  Therefore, for each block within the patch, a histogram of the gradient orientations, 8 circular orientations, are calculated.  This histogram is concatenated for all the 16 blocks and a feature is thus formed:

@@

\noindent with $i$ and $j$ varying belonging to $I = \{ 0,1,2,3 \} $ localizing the 16 blocks within the grid. The angles $\theta$ that belong to $\Theta$ are the eight possible equidistant values between $0$ and $315$.  This vector is normalized, clamped to $0.2$, and re-normalized again.   Details of the method can be found on \citep{Ramele2016,Lowe2004}.  It was implemented using the VLFeat  \citep{Vedaldi2010} public Computer Vision libraries.

@@

Farwell and Donchin P300 Speller \citep{Farwell1988,Rakotomamonjy2008} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.  This procedure exploits a cognitive phenomena raised by the oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.  This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.

The objective of the following experiments is to assess the performance of the algorithms that aim to recognize the shape of the P300 waveform, obtained after averaging signal segments.  This performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.

The problem with any P300 experiment is that users are part of a closed loop with the BCI system and the human is typically considered as a static compliant entity that always performs what the experimenter asks for in a precise and consistent way.  Therefore, P300 experiments are often mined with null-signals.  These are EEG streams which are marked as hit but, because the subject was not particularly focused, or concentrated or so on, the real signal is not generated.   This lack of certainty may be in detriment of analysis of algorithms and can be misleading.   Similar experiments had address this same issue, particularly when benchmarking different algorithms.

@@

Second, an active-modality is also used, where a P300-Based BCI experiment is performed on four subjects, but the signal segments are modified to produce the desired P300 response, but based on information extracted from the same subject.

@@

\begin{itemize}
\item Experiment 1 - Letter Identification Performance: the letter identification performance of each one of these methods on the artificially generated pseudo-real dataset.
\item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERP over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.  The lagging value is picked from a uniform distribution $U(0,0.4)$ [s] ranging from $0$ to $0.4$ of the $1s$ segment size~\citep{DaPelo2018}.
\item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b component of the ERP template is randomly altered.  This component is defined to be located from the stimulus onset between $148$ ms up to $996$ ms which is around $840$ ms long.  This waveform element, multiplied by a gain factor, is subtracted from the original template.  This gain factor between $0$ and $1$ is drawn from a uniform distribution $U(0,1)$.  Additionally this subtracted waveform is multiplied by a Gaussian window with a support of the same length~\citep{Harris1978}.  This avoids adding any discontinuity into the artificial generated signal.
\end{itemize}

@@

\subsubsection{Pseudo-Real Dataset Generation}

\paragraph{Passive Modality}

An EEG stream with null-P300 signal is obtained by the following procedure: 
Four (4) participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and they agree and sign a written informed consent.  This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.  The participants are healthy and have normal or corrected-to-normal vision and no history of neurological disorders. These voluntary subjects are aged between 20-40 years old.  EEG data is collected in a single recording session. Each subject is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of her/him.  The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.  Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10-20 international system.  Reference is set to the right ear lobe and ground is preset as the AFz position.   Sampling frequency is set to 250 Hz.

All participants are instructed to passively watch the flashing screen while not focusing on any particular letter.  None of them have any experience with a BCI device.  A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.

@@

\paragraph{Active modality}

The experiment conducted by~\citep{Riccio2013} is replicated on 4 healthy subjects, performing a copy-spelling task with feedback.  In this case, to construct the pseudo-real dataset each template must first be obtained from each subject.  The most prototypical P300 waveform template is selected by visual observation.  This template is superimposed to each segment marked as \textit{True} on the EEG stream.  The baseline of each segment, on the other hand, is obtained by picking randomly any \textit{False} segment for the same subject.

>>>>

\begin{itemize}
\item Attenuation: Also called suppression or depression. Reduction of amplitude of EEG activity resulting from decreased voltage. When activity is attenuated by stimulation, it is said to have been "blocked" or to show "blocking".
\item Hypersynchrony: Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range. The term suggest an increase in the number of neural elements contributing to the rhythm, or in the synchronization of different neurons with the same discharge pattern~\citep{Buzsaki2012}.
\item Paroxysmal: Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity. 
\item Monomorphic: Activity appearing to be composed of one dominant waveform pattern.
\item Polymorphic: Activity composed of multiple frequencies that combine to form a complex waveform.
\item Transient/Component: An isolated wave or pattern that is distinctly different from background activity. 
\end{itemize}

@@

The traditional clinical approach consists in analyzing the paper strip that is generated by the plot of the signal obtained from the device.  Expert technician and physicians analyze visually the plots looking for specific patterns that may give a hint of the underlying cognitive process or pathology.   Atlases and guidelines were created in order to help in the recognition of these complex patterns.   Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .  The clinical EEG research has also focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements}~\citep{Schomer2010}.

Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)~\citep{Boostani2017,Rodenbeck2006}. The different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines~\citep{Dimitriadis2018}. Visual characterization includes the identification or classification of certain waveform components based on a subjective characterization (e.g. positive or negative peak polarity) or the location within the strip.  It is regular to establish an amplitude difference between different waveforms from which a relation between them is reckoned and a structured index is created (e.g. sleep K-Complex is well characterized based on rates between positive vs negative amplitude)~\citep{Uchida1999}.  Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves, sleep spindles, polysplindles, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).

Moreover, EEG data acquisition is a key procedure during the assessment of patients with focal Epilepsy for potential seizure surgery, where the source of the seizure activity must be reliably identified. The onset of the Epileptic Seizure is defined as the first electrical change seen in the EEG rhythm which can be visually identified from the context and it is verified against any clinical sign indicating seizure onset.  The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges~\citep{EEGIntro}.

Waveform characterization is the method in which analysis has been performed for Event Related Potentials (ERP).  These are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.  ERPs are regularly used to assess auditory response in infants.  They are extensively used and studied in Cognitive Neuroscience~\citep{Luck2005}.  ERPs are identified by their components which are recognizable signal shapes assigned to the observed waveform, that can be linked to some cognitive or measurable psychological process.   One of the most studied ERP is the P300, discovered in 1965 by Sutton, Braren, Zubin and John.  This component is a positive deflection of a subject's EEG signal that arises when an unexpected and infrequent stimulus appears~\citep{Wolpaw2002}.  The P300 is widely utilized in BCI because it can be harnessed to implement a Speller application.  Hence, P300 ERPs are a target phenomena to study by automatic waveform recognition methods.

Table~\ref{tab:methods} summarize a list of depictions used to describe waveforms in the surveyed literature.  Epilepsy has been described by the nature of oscillatory characterization of their waves, like ripples and wiggles, imitating sawtooths or by their geometric shape.  For ERPs on the other hand, more elaborate indexes has been provided, establishing relations between amplitudes of signal components.  Finally, Sleep studies and ICU research are areas where the most complex indexes has been derived, particularly the coupling of signal properties like phase, amplitude and frequency.

@@

The exploration of methods based on waveforms is conducted by following the PRISMA~\citep{Moher2009} guidelines.  Search is performed on Google Scholar, Semantic Web and IEEE Xplore search engines by the terms "Waveforms" OR "Shape" OR "Morphology" OR "Visual inspection" + "EEG".

@@

A related method is used in \citep{Alvarado-Gonzalez2016} where the area under the curve of the EEG is sumarized to derive a feature.  This was even used in the seminal work of Farwell and Donchin on the P300 Speller~\citep{Farwell1988,WolpawJonathanR2012}. Additionally, a logarithmic graph of the peak-to-peak amplitude which is called amplitude integrated EEG (aEEG) \citep{Shah2015} is used nowadays in Neonatal Intensive Care Units.

Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms. In \cite{Klein1976} a feature derived from the amplitude and frequency of its signal and its derivative in time-domain is used.  Moreover, \citep{Yamaguchi2009} explored the use of Mathematical Morphology, where the time-domain structure of contractions and dilations were studied. Finally the proposals of Burch, Fujimori, Uchida and the Period Amplitude Analysis (PAA)~\citep{Uchida1996} algorithm are few of the earliest depictions where the idea of capturing the shape of the signal were established.

@@

Once the plot is generated, its center is used to localize the center of the SIFT patch. This region of the image, where the signal's most important salient waveform should be located, is divided in a grid of $4 x 4$ block and the bidimensional gradient vectors are calculated on each one of them.  Therefore, for each block $(i,j)$ within the patch, a histogram $h(i,j,\theta)$ of the gradient orientations, for 8 circular orientations $\theta$, are calculated.  This histogram is concatenated for all the 16 blocks and a feature is thus formed:

@@

\noindent with $i$ and $j$ belonging to $I = \{ 0,1,2,3 \} $ and localizing the 16 blocks within the grid. The angles $\theta$ that belong to $\Theta$ are the eight possible equidistant values between $0$ and $315$.  This vector is normalized, clamped to $0.2$, and re-normalized again.   Details of the method can be found on \citep{Ramele2016,Lowe2004}.  It was implemented using the VLFeat  \citep{Vedaldi2010} public Computer Vision libraries.

@@

\begin{itemize}
\item EEG Stream: the raw EEG streams for both modalities
\item Templates: P300 templates obtained either from the Subject Number 8 or from the same subject.
\item Pseudo-real dataset
\item null-signals
\end{itemize}

The objective of the following experiments is to assess the performance of the algorithms that aim to recognize the shape of the P300 waveform, obtained after averaging signal segments.  This performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.  The experiments are performed by offline evaluation of the performance on a P300-Based BCI Speller application.

Farwell and Donchin P300 Speller~\citep{Farwell1988,Rakotomamonjy2008} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.  This procedure exploits a cognitive phenomena raised by the oddball paradigm~\citep{Luck2005}: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.  This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.

The problem with any P300 experiment is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.  Therefore, P300 experiments are often mined with \textit{null-signals}.  These are EEG streams which are marked as hit but, because the subject was not particularly focused, or concentrated or so on, and the real signal is not generated.   This lack of certainty may be in detriment of analysis of algorithms and can be misleading or difficult to dealt with.   Similar experiments had address this same issue, particularly when benchmarking different algorithms.

\subsubsection{Passive Modality}

@@

After the modification it was verified that the flickering effect of the 4 Hz signal is mantained and that alpha

\subsubsection{Active Modality}

Second, an active-modality is also used, where a P300-Based BCI experiment is performed on four subjects, but the signal segments are modified to produce the desired P300 response, but based on information extracted from the same subject.
In this case, to construct the pseudo-real dataset each template must first be obtained from each subject.  The most prototypical P300 waveform template is selected by visual observation.  This template is superimposed to each segment marked as \textit{True} on the EEG stream.  The baseline of each segment, on the other hand, is obtained by picking randomly any \textit{False} segment for the same subject.

@@

\begin{itemize}
\item Experiment 1 - Letter Identification Performance: the letter identification performance of each one of these methods on the artificially generated pseudo-real dataset.  The pool of $70$ P300 ERP waveforms, either obtained from the same subject in the passive-modality or from each subject in the active-modality are used to seed the EEG streams.  The templates are randomly selected.
\item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERPs over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.  The lagging value is picked from a uniform distribution $U(0,0.4)$ [s] ranging from $0$ to $0.4$ of the $1s$ segment size~\citep{DaPelo2018}.
\item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b component of the ERP template is randomly altered.  This component is defined to be located from the stimulus onset between $148$ ms up to $996$ ms which is around $840$ ms long.  This waveform element, multiplied by a gain factor, is subtracted from the original template.  This gain factor between $0$ and $1$ is drawn from a uniform distribution $U(0,1)$.  Additionally this subtracted waveform is multiplied by a Gaussian window with a support of the same length~\citep{Harris1978}.  This avoids adding any discontinuity into the artificial generated signal.
\end{itemize}

@@

\subsubsection{EEG Stream Generation}

An EEG stream with null-P300 signal is obtained by the following procedure:

Eight (8) healthy participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and they agree and sign a written informed consent.  This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.  The participants are healthy and have normal or corrected-to-normal vision and no history of neurological disorders. These voluntary subjects are aged between 20-40 years old.  EEG data is collected in a single recording session. Each subject is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of her/him.  The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.  Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10-20 international system.  Reference is set to the right ear lobe and ground is preset as the AFz position.   Sampling frequency is set to 250 Hz.

The experimental protocol is composed of $35$ trials to spell $7$ words of $5$ letters each.  Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.  This yields exactly $120$ intensifications of rows and columns per trial.  The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ seconds.  This provides a $4$ Hz frequency of flashes on the screen.  The initial pause and the inter-trial pauses are set to $20$ seconds.  The whole experiment lasts for around $1400$ seconds.

Four out of the eight participants are instructed to passively watch the flashing screen while not focusing on any particular letter. They do not receive any extra information on the screen.  None of them have any experience with a BCI device.  A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.

The remaining four participants perform a copy-spelling task where the computer monitor suggests by highlighting the target letter, which letter the subject needs to focus.  During each selection letter trial, the target letter is suggested at the bottom of the screen.

@@
